```{r setup, echo = FALSE, warning = FALSE, message = FALSE}

library(knitr)
library(dplyr)
library(huxtable)
options(
        huxtable.knit_print_df       = FALSE, 
        huxtable.add_colnames        = TRUE,  # needed when run by testthat
        huxtable.latex_use_fontspec  = TRUE,
        huxtable.bookdown            = TRUE
      )

is_latex <- guess_knitr_output_format() == "latex"
# is_latex <- TRUE
knitr::knit_hooks$set(
  barrier = function(before, options, envir) {
    if (! before && is_latex) knitr::asis_output("\\FloatBarrier")
  }
)
```

```{r, figsetup5, include=FALSE}
knitr::opts_chunk$set(
  fig.align = 'center',
  out.width = '80%'
)
```

```{r, echo = FALSE}
include_cropped_graphics <- function(x) {
  knitr::include_graphics(knitr::plot_crop(x))
#  knitr::include_graphics(x)
}
```

# Simple distributions {#simple}

In this section we use a dataset to introduce some commonly-used simple distributions. We will study the **discrete** distributions: Bernoulli, binomial, geometric and Poisson. We will also study the **continuous** distributions: uniform, exponential, normal.

## The Bernoulli distribution

## The binomial distribution {#binomial}

## The geometric distribution

## The Poisson distribution

## The uniform distribution

## The exponential distribution

## The normal distribution {#normal}

## QQ plots