--- 
title: "STAT0002 Introduction to Probability and Statistics"
author: "Paul Northrop"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
classoption: openany, a4paper
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Produces STAT0002 notes in an accessible format"
lang: "en-gb"
fontsize: 11pt
---

# The purpose of these notes {-}

These notes functions as a textbook for the material covered in STAT0002. The teaching events in STAT0002 will follow the general order of the topics covered in these notes. Chapters 1 and 2 are covered over Weeks 1 and 2 and from then on we will cover one chapter per teaching week.  There will be a section on Moodle dedicated to each week of term.

See the following sections STAT0002 Moodle page for important information:

* **Module overview**: general information;
* **Asking for help**: how to ask questions;
* **Assessment and feedback**: how you will be assessed and when;
* **Exercises**: (feedback on) weekly exercises sheets and summaries of tutorial material;
* **Quizzes**: Moodle quizzes to help you to check your understanding;
* **stat0002 R package**: provides a link to the [stat0002 R package](https://paulnorthrop.github.io/stat0002/), which contains R-related material to assist your learning;
* **Past exams**: (towards the end of the module) past exam papers and information about exams.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r kntir_auto_pdf, echo = FALSE, warning = FALSE, message = FALSE}
options(knit.graphics.auto_pdf = TRUE)
```

```{r, figsetup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = 'center',
  out.width = '75%'
)
```
